An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-02 rule-ltl.tex
fixing problems

\begin{theorem}
\label{theo1}
The expressive power of 
ELTL is the same as QLTL.
\end{theorem}
\noindent {\bf Proof.}
Each ELTL formula $\eta$, as defined above,
is expressible in QLTL using the following formula:
\[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
\]
For the other direction, one can translate any QLTL property into an $\omega$ automaton\footnote{The past operators of a QLTL property do not change this fact, since the translation can be done first to second order monadic logic, which can be translated into an $\omega$-automaton}.
We assume that the property is given as a deterministic Muller automaton $M = (S, i , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $i \in S$ is the initial state, the set $\Sigma$ consists of the inputs, each represented as a Boolean combination of the variables $B$, $\Delta : S \times \Sigma \mapsto S$, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set, 
where an accepted input needs to pass infinitely often in exactly
one of the specified sets. To formulate an ELTL formula,
he auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.

Then we translate $M$ into an ELTL formula
of the form \begin{tabbing}
$\bigvee_{F \in {\cal F}} (\bigwedge_{s \in F} \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
where \\
\ \ \ \ \ $a_s := \vee _{\Delta ( s'
, \alpha ) = s } 
\alpha \wedge \ominus a_{s'} )
\wedge (s=i \rightarrow \neg \ominus \true)
\, :\,  s \in S
$
\end{tabbing}
With $(s=i)$ is $\true$ when $s$ is $i$, and $\false$ otherwise.
Basically, the assignments to the auxiliary variables are used to simulate the change of states\footnote{It is possible to use only $\lceil \log ( | S | ) \rceil$
variables, at the expense of longer and more complicated formulas.}, and the main formula expresses the Muller acceptance condition. \qed

The use of the auxiliary variables $a_1 \ldots a_n$ in the
rules of an ELTL specification 
$\eta$ can be viewed as representing a
state machine, in particular, in
particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an ELTL specification $\eta$ as consisting with two synchronous parts: an LTL property 
(the statement) $\psi$
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to QLTL.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

@@

For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
\begin{lemma}
\label{simplecase}
Any safety QPLTL property can be expressed using the following restricted
ELTL form:
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}  \] 
 with $p$ being one of the auxiliary variables $a_j$.
\end{lemma}
\noindent {\bf Proof.} The proof of the Lemma follows from the
axiomatization of QLTL in~\cite{MP}, e.g.,
$\varphi \since \psi \leftrightarrow ( \psi \vee ( \varphi \wedge \varphi \since \psi) ) )$. \qed

@@

Runtime verification of temporal temporal specification often
concentrates on the past portion of the logic.  Past specifications have the important property that one can distinguish when they are violated. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.

@@

The runtime verification exploits the fact that
the semantic definition is recursive in both the structure of the property, where subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, operates on two vectors (arrays) of values indexed by subformulas:  $\old$ that summarizes the truth values of the
subformulas just {\em before} the new state, and $\current$ for the {\em current} state.

@@

Runtime verification of QPLTL can be done by translating a
formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

The well foundedness of the formulas defining the new auxiliary variables in EPLTL makes the extension of the above
runtime algorithm simple.
In order to extend the algorithm to perform runtime verification of an EPLTL formula $\psi$, as defined in~(\ref{ELTL}),  we need to  add to the above
algorithm memory for $\current ( a_j )$ and $\old ( a_j )$
and perform, between steps 3 and 4, the following, per each $a_j$: \\

\ \ \ \ -\ $\current ( a_j ) = \current ( \varphi_j )$ \\

>>>>

\begin{theorem}
\label{theo1}
The expressive power of 
ELTL is the same as QLTL.
\end{theorem}
\noindent {\bf Proof.}
Each ELTL formula $\eta$, as defined above,
is expressible in QLTL using the following formula:

\[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
\]

\klaus{How do we know that the ``other direction'' below is possible?}

For the other direction, one can translate any QLTL property into an $\omega$ automaton\footnote{The past operators of a QLTL property do not change this fact, since the translation can be done first to second order monadic logic, which can be translated into an $\omega$-automaton.}.
We assume that the property is given as a deterministic Muller automaton $M = (S, i , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $i \in S$ is the initial state, $\Sigma$ is the set of inputs, each represented as a Boolean combination of the variables $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set, 
where an accepted input needs to pass infinitely often in exactly
one of the specified sets. To formulate an ELTL formula,
he auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.

\klaus{The following formulation above is not not clear: ``where an accepted input needs to pass infinitely often in exactly
one of the specified sets''.}

\noindent
Then we translate $M$ into an ELTL formula
of the form:

\begin{tabbing}
\ \ \ \ \ $\bigvee_{F \in {\cal F}} (\bigwedge_{s \in F} \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
where \\
\ \ \ \ \ \ \ \ \ \ $a_s := \vee _{\Delta ( s'
, \alpha ) = s } 
\alpha \wedge \ominus a_{s'} )
\wedge (s=i \rightarrow \neg \ominus \true)
\, :\,  s \in S
$
\end{tabbing}

\klaus{Above: should $s=i \rightarrow \neg \ominus \true$ instead be (note: two changes):
$\neg \ominus \true \rightarrow a_s=i$
}

\klaus{Above: can't judge whether the $\bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$ is correct.}

\noindent
With $(s=i)$ is $\true$ when $s$ is $i$, and $\false$ otherwise.
Basically, the assignments to the auxiliary variables are used to simulate the change of states\footnote{It is possible to use only $\lceil \log ( | S | ) \rceil$
variables, at the expense of longer and more complicated formulas.}, and the main formula expresses the Muller acceptance condition. \qed

The use of the auxiliary variables $a_1 \ldots a_n$ in the
rules of an ELTL specification 
$\eta$ can be viewed as representing a
state machine, in particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an ELTL specification $\eta$ as consisting with two synchronous parts: an LTL property 
(the statement) $\psi$
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to QLTL.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}

@@

\klaus{Last sentence above: perhaps say with words that we are always in an acceptance state - if that's what the point is.}

For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
\begin{lemma}
\label{simplecase}
Any safety QPLTL property can be expressed using the following restricted
ELTL form:
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}  \] 
 with $p$ being one of the auxiliary variables $a_j$.
\end{lemma}
\noindent {\bf Proof.} The proof of the Lemma follows from the
axiomatization of QLTL in~\cite{MP}, e.g.,
$\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed

\klaus{In the axiom above, isn't there missing a $\ominus$ before
the second $(\varphi\ \since\ \psi)$?}

\klaus{Above: is the following really correct?: 
``with $p$ being one of the auxiliary variables $a_j$''.}

@@

Runtime verification of temporal specifications often
concentrates on the past portion of the logic.  Past-time  specifications have the important property that one can distinguish when they are violated. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.

@@

\noindent
Runtime verification exploits the fact that
the semantic definition is recursive in both the structure of the property, where subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, operates on two vectors (arrays) of values indexed by subformulas:  $\old$ that summarizes the truth values of the
subformulas just {\em before} the new state, and $\current$ for the {\em current} state.

@@

\noindent
Runtime verification of QPLTL can be done by translating a
formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. 
However, QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

The well foundedness of the formulas defining the new auxiliary variables in EPLTL makes the extension of the above
runtime algorithm simple.
In order to extend the algorithm to perform runtime verification of an EPLTL formula $\psi$, as defined in~(\ref{ELTL}),  we need to  add to the above
algorithm memory for $\current ( a_j )$ and $\old ( a_j )$,
and perform the following between steps 3 and 4 for each $a_j$: \\

\ \ \ \ \ $\current ( a_j ) := \current ( \varphi_j )$ \\
